---
title: "Getting Started with freestiler"
vignette: >
  %\VignetteIndexEntry{Getting Started with freestiler}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| eval: false
library(freestiler)
```

## Overview

freestiler is a Rust-powered vector tile engine for R. It takes sf data frames
and produces [PMTiles](https://github.com/protomaps/PMTiles) archives --- self-contained, single-file
tilesets that can be served statically or viewed locally. The entire pipeline
runs in-memory with zero external dependencies: no tippecanoe, no Java, no Go.

freestiler supports two tile formats:

- **MLT (MapLibre Tiles)** --- the default. A next-generation columnar binary
  format for efficient vector tile encoding.
- **MVT (Mapbox Vector Tiles)** --- the widely supported protobuf-based format.

## Installation

Install from [r-universe](https://walkerke.r-universe.dev):

```r
install.packages("freestiler", repos = "https://walkerke.r-universe.dev")
```

Or install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("walkerke/freestiler")
```

## Creating your first tileset

The main function is `freestile()`. Pass it an sf data frame and an output path:

```{r}
#| eval: false
library(sf)
library(freestiler)

nc <- st_read(system.file("shape/nc.shp", package = "sf"))

freestile(nc, "nc_counties.pmtiles", layer_name = "counties")
```

```
Creating MLT tiles (zoom 0-14) for 100 features across 1 layer...
  Tiling layer 'counties' (zoom 0-14)...
Created nc_counties.pmtiles (65.2 KB)
```

This creates a PMTiles archive with MLT-encoded tiles at zoom levels 0 through 14.

## Viewing tiles with mapgl

Use the [mapgl](https://walker-data.com/mapgl/) package to view your tileset
directly in R:

```{r}
#| eval: false
library(mapgl)

maplibre() |>
  add_vector_source(
    id = "counties",
    url = paste0("pmtiles://", normalizePath("nc_counties.pmtiles"))
  ) |>
  add_fill_layer(
    id = "county-fill",
    source = "counties",
    source_layer = "counties",
    fill_color = "#00897b",
    fill_opacity = 0.5
  ) |>
  add_line_layer(
    id = "county-borders",
    source = "counties",
    source_layer = "counties",
    line_color = "#004d40",
    line_width = 1
  )
```

## MLT vs MVT format

The default format is MLT (MapLibre Tiles). To use MVT instead:
```{r}
#| eval: false
freestile(nc, "nc_mvt.pmtiles", layer_name = "counties", tile_format = "mvt")
```

MLT is a columnar encoding that can be more compact than MVT, especially for
polygon-heavy datasets. MVT is the universal fallback supported by all mapping
libraries. Use MVT when maximum compatibility is needed.

## Controlling zoom levels

The `min_zoom` and `max_zoom` parameters control the zoom range:

```{r}
#| eval: false
freestile(nc, "nc_z4_10.pmtiles",
  layer_name = "counties",
  min_zoom = 4,
  max_zoom = 10
)
```

## Feature dropping with drop_rate

For large datasets, `drop_rate` provides exponential feature thinning at lower
zoom levels. Points are thinned using spatial ordering (Morton curve) to
maintain even coverage; polygons and lines are thinned by area:

```{r}
#| eval: false
freestile(nc, "nc_dropping.pmtiles",
  layer_name = "counties",
  drop_rate = 2.5
)
```

The `base_zoom` parameter controls the zoom level above which all features are
present. Below `base_zoom`, features are progressively thinned. By default,
`base_zoom` equals `max_zoom`:

```{r}
#| eval: false
freestile(nc, "nc_basezoom.pmtiles",
  layer_name = "counties",
  drop_rate = 2.5,
  base_zoom = 10
)
```

## Multi-layer tiles

Pass a named list to create multi-layer tilesets:

```{r}
#| eval: false
pts <- st_centroid(nc)

freestile(
  list(counties = nc, centroids = pts),
  "nc_layers.pmtiles"
)
```

Use `freestile_layer()` for per-layer zoom control:

```{r}
#| eval: false
freestile(
  list(
    counties = freestile_layer(nc, min_zoom = 0, max_zoom = 10),
    centroids = freestile_layer(pts, min_zoom = 6, max_zoom = 14)
  ),
  "nc_layers.pmtiles"
)
```

## Point clustering

For point layers, `cluster_distance` merges nearby points into clusters with
a `point_count` attribute:

```{r}
#| eval: false
freestile(pts, "nc_clustered.pmtiles",
  layer_name = "centroids",
  cluster_distance = 50,
  cluster_maxzoom = 8
)
```

## Feature coalescing

The `coalesce` parameter merges features with identical attributes within each
tile. Lines sharing endpoints are joined; polygons are grouped into
MultiPolygons:

```{r}
#| eval: false
freestile(nc, "nc_coalesced.pmtiles",
  layer_name = "counties",
  coalesce = TRUE
)
```
